Probabilistic Catalogs for Crowded Stellar Fields

### 2013-01-18 paper/ms.tex
Fixed variance
Thank referee
Hobson reference
New parameter values for new simulated datasets
Ranges wrong

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if an unknown-$N$ model had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods have also been used to model crowded fields
\citep[e.g. maximum likelihood,][]{irwin}.
However, maximum likelihood is not completely appropriate in flexible models
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise. Various other techniques have
also been proposed in the literature \citep[e.g.][]{2002MNRAS.335...73M, 2007A&A...461..373M, 2009PASP..121.1429Z}.

@@

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times n$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N,
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N f_i \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of each pixel is determined by the brightness of the sky and the
brightness of the mock image in that pixel:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta I_{ij}).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

\subsection{The Prior Distribution}
The prior distribution for the number of stars $N$ is assigned to be uniform
between 0 and some maximum number $N_{\rm max}$. The extent of the image is
assumed to be from $x=x_{\rm min}$ to $x=x_{\rm max}=x_{\rm min} + x_{\rm range}$ and from
$y=y_{\rm min}$ to $y=y_{\rm max}=y_{\rm min} + y_{\rm range}$ in arbitrary units, and the
positions of the stars are assigned independent uniform priors:
\begin{eqnarray}
x_i &\sim& \textnormal{Uniform}(x_{\rm min} - 0.05x_{\rm range}, x_{\rm max} + 0.05x_{\rm range}) \\
y_i &\sim& \textnormal{Uniform}(y_{\rm min} - 0.05y_{\rm range}, y_{\rm max} + 0.05y_{\rm range})
\end{eqnarray}
The stars are allowed to be slightly outside of the observed image because the
PSF can scatter light from these stars into the image. Our model for the
luminosity function is a broken power-law with four
free parameters:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
where $h_1$ is a lower flux limit, $h_2$ is a break-point, $\alpha_1$ is
the slope of the distribution between $h_1$ and $h_2$, and $\alpha_2$ is the
slope of the distribution above $h_2$. For details on the broken power-law
model, see Appendix~\ref{power_law}. The prior distribution on $h_1$, $h_2$,
$\alpha_1$, and $\alpha_2$ is assigned to be:
\begin{eqnarray}
\ln h_1 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^3)) \\
\ln h_2 &\sim& \textnormal{Uniform}(\ln(h_1), \ln(h_1) + 2.3) \\
\alpha_1 &\sim& \textnormal{Uniform}(1, 5) \\
\alpha_2 &\sim& \textnormal{Uniform}(1, 5).
\end{eqnarray}
These priors express vague prior knowledge about $\alpha_1$ and $\alpha_2$ in
addition to vague prior knowledge about $h_1$ and $h_2$ apart from the fact
that the flux units are not extreme and that $h_2$ should be no more than an
order of magnitude greater than $h_1$.

@@

\begin{table}\footnotesize
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value (Test Case 1) & Value (Test Case 2)\\
\hline
$N$ & 100 & 1000\\
$h_1$ & 0.3 & 0.3\\
$h_2$ & 0.6 & 0.6\\
$\alpha_1$ & 1.1 & 1.1\\
$\alpha_2$ & 2 & 2\\
\hline
$\sigma_0$ & 0.05 & 0.3\\
$\eta$ & 0 & 0 \\
$s_1$ & 0.02 & 0.02\\
$s_2$ & 0.1 & 0.1\\
$w$ & 0.5 & 0.5\\
\hline
\end{tabular}
\end{center}
\caption{True parameter values used to generate the simulated data. The main
difference between the two test cases is the number of stars and the noise
level, both of which are higher in test case 2.
\label{tab:truth}}
\end{table}

@@

\section{Acknowledgements}
It is a pleasure to thank
  Jonathan Goodman (NYU),
  Fengji Hou (NYU),
  Dustin Lang (CMU),
  Geraint Lewis (Sydney), and
  Phil Marshall (Oxford)
for their comments and discussions.
BJB would like to thank Tommaso Treu (UCSB) for his support, and
Wayne Stewart and Arden Miller (Auckland) for their encouragement and advice.
DFM and DWH were partially supported by
   the US National Science Foundation (grant IIS-1124794) and
   the US National Aeronautics and Space Administration (grant NNX12AI50G).

@@

>>>>

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2003MNRAS.338..765H, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if an unknown-$N$ model had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods have also been used to model crowded fields
\citep[e.g. maximum likelihood,][]{irwin}.
However, maximum likelihood is not completely appropriate in flexible models
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise. Various other techniques have
also been proposed in the literature \citep[e.g.][]{2002MNRAS.335...73M, 2007A&A...461..373M, 2009PASP..121.1429Z}.

@@

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times n$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N,
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N f_i \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of each pixel is determined by the brightness of the sky and the
brightness of the mock image in that pixel:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

\subsection{The Prior Distribution}
The prior distribution for the number of stars $N$ is assigned to be uniform
between 0 and some maximum number $N_{\rm max}$. The extent of the image is
assumed to be from $x=x_{\rm min}$ to $x=x_{\rm max}=x_{\rm min} + x_{\rm range}$ and from
$y=y_{\rm min}$ to $y=y_{\rm max}=y_{\rm min} + y_{\rm range}$ in arbitrary units, and the
positions of the stars are assigned independent uniform priors:
\begin{eqnarray}
x_i &\sim& \textnormal{Uniform}(x_{\rm min} - 0.1x_{\rm range}, x_{\rm max} + 0.1x_{\rm range}) \\
y_i &\sim& \textnormal{Uniform}(y_{\rm min} - 0.1y_{\rm range}, y_{\rm max} + 0.1y_{\rm range})
\end{eqnarray}
The stars are allowed to be slightly outside of the observed image because the
PSF can scatter light from these stars into the image. Our model for the
luminosity function is a broken power-law with four
free parameters:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
where $h_1$ is a lower flux limit, $h_2$ is a break-point, $\alpha_1$ is
the slope of the distribution between $h_1$ and $h_2$, and $\alpha_2$ is the
slope of the distribution above $h_2$. For details on the broken power-law
model, see Appendix~\ref{power_law}. The prior distribution on $h_1$, $h_2$,
$\alpha_1$, and $\alpha_2$ is assigned to be:
\begin{eqnarray}
\ln h_1 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^3)) \\
\ln h_2 &\sim& \textnormal{Uniform}(\ln(h_1), \ln(h_1) + 2.3) \\
\alpha_1 &\sim& \textnormal{Uniform}(1, 5) \\
\alpha_2 &\sim& \textnormal{Uniform}(1, 5).
\end{eqnarray}
These priors express vague prior knowledge about $\alpha_1$ and $\alpha_2$ in
addition to vague prior knowledge about $h_1$ and $h_2$ apart from the fact
that the flux units are not extreme and that $h_2$ should be no more than an
order of magnitude greater than $h_1$.

@@

\begin{table}\footnotesize
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value (Test Case 1) & Value (Test Case 2)\\
\hline
$N$ & 100 & 1000\\
$h_1$ & 0.3 & 0.3\\
$h_2$ & 0.6 & 0.6\\
$\alpha_1$ & 1.1 & 1.1\\
$\alpha_2$ & 2 & 2\\
\hline
$\sigma_0$ & 10 & 10\\
$\eta$ & 10 & 10 \\
$s_1$ & 0.02 & 0.02\\
$s_2$ & 0.1 & 0.1\\
$w$ & 0.5 & 0.5\\
\hline
\end{tabular}
\end{center}
\caption{True parameter values used to generate the simulated data. The main
difference between the two test cases is the number of stars and the noise
level, both of which are higher in test case 2.
\label{tab:truth}}
\end{table}

@@

\section{Acknowledgements}
It is a pleasure to thank
  Jonathan Goodman (NYU),
  Fengji Hou (NYU),
  Dustin Lang (CMU),
  Geraint Lewis (Sydney), and
  Phil Marshall (Oxford)
for their comments and discussions.
The referee is thanked for their thoughtful comments which helped us to improve
the paper.
BJB would like to thank Tommaso Treu (UCSB) for his support, and
Wayne Stewart and Arden Miller (Auckland) for their encouragement and advice.
DFM and DWH were partially supported by
   the US National Science Foundation (grant IIS-1124794) and
   the US National Aeronautics and Space Administration (grant NNX12AI50G).

@@

\bibitem[Hobson 
\& McLachlan(2003)]{2003MNRAS.338..765H} Hobson, M.~P., \& McLachlan, C.\ 2003, \mnras, 338, 765
